---
title: "Environmental Data Demo (R Integration)"
format:
  html:
    df-print: paged
---

This page demonstrates the computational requirement of using R code within a Quarto website, using a starter environmental dataset.

## Step 1: Load the dataset

```{r}
env <- read.csv("data/environment_nz.csv")
head(env)
```

## Step 2: Quick summary values

```{r}
avg_pm25 <- mean(env$pm25)
total_rainfall <- sum(env$rainfall_mm)
avg_nitrate <- mean(env$river_nitrate_mgL)

data.frame(
  metric = c("Average PM2.5", "Total Rainfall (mm)", "Average River Nitrate (mg/L)"),
  value = c(round(avg_pm25, 2), round(total_rainfall, 1), round(avg_nitrate, 2))
)
```

## Step 3: Rainfall by month (bar chart)

```{r}
barplot(
  env$rainfall_mm,
  names.arg = env$month,
  main = "Monthly Rainfall",
  xlab = "Month",
  ylab = "Rainfall (mm)",
  las = 2
)
```

## Step 4: PM2.5 and river nitrate trend

```{r}
plot(
  env$month,
  env$pm25,
  type = "o",
  col = "steelblue",
  ylim = c(min(c(env$pm25, env$river_nitrate_mgL)), max(c(env$pm25, env$river_nitrate_mgL))),
  xlab = "Month",
  ylab = "Value",
  main = "PM2.5 and River Nitrate Across the Year"
)
lines(env$month, env$river_nitrate_mgL, type = "o", col = "darkgreen")
legend("topright", legend = c("PM2.5", "River Nitrate"), col = c("steelblue", "darkgreen"), lty = 1)
```

## Interpretation prompt

Write 2–3 sentences under each graph explaining:

- what pattern the graph shows,
- one possible reason for the trend,
- one limitation of the dataset.

## Option B demo: Waste and recycling dataset

```{r}
waste <- read.csv("data/waste_recycling_nz.csv")
head(waste)
```

```{r}
plot(
  waste$month,
  waste$landfill_tonnes,
  type = "o",
  col = "firebrick",
  ylim = c(min(c(waste$landfill_tonnes, waste$recycling_tonnes)), max(c(waste$landfill_tonnes, waste$recycling_tonnes))),
  xlab = "Month",
  ylab = "Tonnes",
  main = "Landfill vs Recycling Across the Year"
)
lines(waste$month, waste$recycling_tonnes, type = "o", col = "darkcyan")
legend("topright", legend = c("Landfill", "Recycling"), col = c("firebrick", "darkcyan"), lty = 1)
```
